---
title: "Parallel Computing with coMethDMR"
author: "Gabriel J. Odom, PhD, ThD"
date: "7/16/2019"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example Data
We will use the PFC Data from Lunnon et al. (2014).



# Serial Computing
First we load the packages we will need.
**TO DO: comment on why we need the `DMRcate` package.**
```{r packLoad, message=FALSE}
library(coMethDMR)
library(DMRcate)
```

All of these computations were performed on a Dell Precision 5810 with 64Gb of RAM, an Intel Xeon E5-2640 CPU at 2.40Ghz, and using up to 20 cores.

## Dasen Normalized Data
**TO DO: add source and cleaning scripts for `pfc_df` and `pfcPheno_df`.**
At current, these data sets are saved locally. The original data sets are available online, but we have pre-processed these data sets. We are working to make the import and cleaning scripts for these data sets publically available.
```{r data}
projectPath_char <-
  "~/Dropbox (BBSR)/GabrielOdom/coMethDMR/vignette_parallel_computing/"
pfc_df <- readRDS(
  paste0(projectPath_char, "pfc_df.RDS")
)
pfcPheno_df <- readRDS(
  paste0(projectPath_char, "pfcPheno_df.RDS")
)
```

The PFC data is a data frame of methylation *beta* values with CpG locations in the rows and samples in the columns.
```{r inspectPFC}
dim(pfc_df)
pfc_df[1:5, 1:5]
```

The phenotype data is a data frame with response (e.g. `stage`) and covariate information (e.g. `age.brain` and `sex`) for the given subjects.
```{r inspectPheno}
str(pfcPheno_df)
```

## List of Island Regions
Because we are interested in regions of the genome that are concurrently methylated, we need a list of CpG probe IDs that belong to our region of choice. The file `"ISLAND3_200.rds"` holds a list of all CG locations that fit the following criteria: they are found in CpG Island regions, the maximum distance between consecutive probes is 200 base pairs, and there are at least 3 CpGs in the cluster.
```{r importRegions}
closeByGenomicRegion_ls <- readRDS(
  system.file(
    "extdata",
    "ISLAND3_200.rds",
    package = 'coMethDMR',
    mustWork = TRUE
  )
)
closeByGenomicRegion_ls <- unname(closeByGenomicRegion_ls)

head(closeByGenomicRegion_ls)
```

If you would like to use a different region, change the file name to match the region of interest (for the gene body region, use `"GENEBODY3_200.rds"`). If you would like to use a different maximum probe distance or different minimum number of CpGs, use the `WriteCloseByAllRegions()` function to compute and save such a list.

## Find All Co-Methylated Regions
The `CoMethAllRegions()` function enables you specify your own list of regions based on a region type (through the `file` or `CpGs_ls` arguments), or to use the pre-computed regions lists (through the `regionType` argument). This function currently takes quite a while to run (about 75 minutes), so don't use it on larger data sets. In fact, this is the motivation for this vignette.
```{r serial_CoMethAllRegions, eval = FALSE}
# DON'T RUN
a <- Sys.time()
pfc_cgi_ls <- CoMethAllRegions(
  betaMatrix = pfc_df,
  regionType = "ISLAND",
  arrayType = "450k",
  rDropThresh_num = 0.4,
  returnAllCpGs = FALSE
)
Sys.time() - a
# 1.262612 hours
```

## Apply the Linear Mixed Model on All Regions
Once we combine the pre-defined region clusters with the methylation data to find the co-methylated regions, we can find which co-methylated regions are related to the chosen output (conditional on the covariates specified).
```{r serial_lmmTestAllRegions, eval = FALSE}
a <- Sys.time()
res_cgi_df1 <- lmmTestAllRegions(
  # methylation data
  beta_df = pfc_df,
  # list of co-methylated regions
  region_ls = pfc_cgi_ls,
  # phenotype data frame
  pheno_df = pfcPheno_df,
  # column of the phenotype data to be used as the response
  contPheno_char = "stage",
  # columns of the phenotype data to be used as covariates
  covariates_char = c("age.brain", "sex", "Mplate", "prop.neuron"),
  # model type
  modelType = "randCoef",
  # Type of probes
  arrayType = "450k"
)
Sys.time() - a
# 34.02621 min for 4892 regions
```

For a single row, this will return a table as shown:

| chrom |    start |      end | nCpGs | Estimate | StdErr |    Stat | pValue |    FDR |
|-------|----------|----------|-------|----------|--------|---------|--------|--------|
| chr22 | 18268062 | 18268249 |     3 |  -0.0732 | 0.0394 | -1.8564 | 0.0634 | 0.0634 |


# Using the **parallel** Package
Load the package. This package should be part of the base packages installed when you first installed R.
```{r parallelPackLoad}
library(parallel)
```

## Cluster Setup
We first need a short discussion on parallel computing in Windows compared to Macintosh or Linux. Linux and MacOS are able to share information across multiple workers, while Windows does not allow sharing. This means that any data necessary for parallel computation must be copied for each worker in a Windows environment. Therefore, for larger problems, you may only be able to use 3-4 workers, even if you have 8 or more available.

### First Comment on Memory
Loading the **coMethDMR**:: package requires 3.5Gb of available RAM. This is in addition to the 0.4Gb of RAM necessary to load the PFC data into memory. This cost is driven by the annotation data sets included in the many database packages that **coMethDMR** depends upon. In the next version of this package, we will remove dependencies on as many of these data sets as possible, and move all necessary annotation data sets to a supplemental data package.

Now we create the cluster of workers. Each worker will need access to the **coMethDMR**:: package, so budget for 3.5Gb of RAM per worker. For a machine with 64Gb of RAM, 12 workers will use most of this. 
```{r parallel_cluster, eval=FALSE}
clust <- makeCluster(12)
```

Load the packages you need on these workers.
```{r parallel_packLoad, eval=FALSE}
clusterEvalQ(cl = clust, library(coMethDMR))
clusterEvalQ(cl = clust, library(DMRcate))
```
**DMRcate**:: only uses 0.1Gb of RAM, so the total RAM for packages is `12 * (3.5 + 0.1) = ` 43.2Gb.

Export the three data pieces to each cluster.
```{r parallel_dataExport, eval=FALSE}
clusterExport(
  cl = clust,
  varlist = c("pfc_df", "pfcPheno_df", "closeByGenomicRegion_ls")
)
```
The data we need is just under 0.5Gb, so we are now up to 48.6Gb of RAM for the 12 workers, and just under 4Gb for the master (52.2Gb total).

## Computing the Co-Methylated Regions in Parallel
Now that we have a master and 12 workers initialized, we can use the **parallel** apply function, `parLapply()`.
```{r parLapply, eval=FALSE}
a <- Sys.time()
coMethCpGsAllRegions_ls <- parLapply(
  # Specify the cluster of workers
  cl = clust,
  # The list of regions to apply over
  X = closeByGenomicRegion_ls,
  # The function to apply
  fun = CoMethSingleRegion,
  # Additional data passed to CoMethSingleRegion()
  betaMatrix = pfc_df,
  method = "pearson",
  arrayType = "450k",
  returnAllCpGs = FALSE
)
Sys.time() - a
# 12.66661 min over 12 cores
```
We have decreased the computing time by a factor of 6 because we used 12 cores. This is an important concept to understand: using 10 cores does not cut the computing time by a factor of 10. While the workers are very helpful, they need the time to start working, they require constant "guidance" and input from the master, and they also need to return their components of the finished product to the master. Then, the master has to compile all of the results from the workers. This takes time. In fact, linearly increasing the number of workers does not linearly decrease the computational time. In fact, increasing from 8 workers to 12 workers only decreased the time from 14.5 to 12.7 minutes (all while increasing the RAM required from 36.3Gb to 52.2Gb). If you only have 32Gb of RAM available, use 5-6 workers (this computation completes in 17.8 minutes for 6 workers).

### Post Compute Wrangling
The first element of this list is a list with the following components.
```{r inspect_lmmTestOut, eval=FALSE}
coMethCpGsAllRegions_ls[[1]]
```
```
$contiguousRegions
                     Region        CpG   Chr   MAPINFO      r_drop keep keep_contiguous
1 chr10:100028236-100028499 cg20214853 chr10 100028236 -0.02236143    0               0
2 chr10:100028236-100028499 cg04677227 chr10 100028320  0.44360382    1               1
3 chr10:100028236-100028499 cg11632906 chr10 100028468  0.52502419    1               1
4 chr10:100028236-100028499 cg07146435 chr10 100028499  0.58816893    1               1

$CpGsSubregions
$CpGsSubregions$`chr10:100028320-100028499`
[1] "cg04677227" "cg11632906" "cg07146435"
```


Because the output of `CoMethSingleRegion()` is a nested list of lists of two elements, and we only care about the individual CpG sub-regions (the `CpGsSubregions` elements), we will extract these elements directly.
```{r post_parallel, eval=FALSE}
pfc_cgi_ls <- unlist(
  lapply(coMethCpGsAllRegions_ls, `[[`, 2),
  recursive = FALSE
)
```

## Computing the Linear Mixed Models in Parallel
Instead of applying the mixed model to an entire data frame, we apply the model to the data subsets matching the co-methylated clusters found in the previous section.

### Creating the Cluster-Specific Data Subsets
First, we create a list of cluster-specific data subsets. Because this is a relatively simple computation, we would probably spend more time setting up a cluster of workers than the 2.5 minutes this computation takes to complete.
```{r list_of_DFs,eval=FALSE}
a0 <- Sys.time()
CpGnames <- rownames(pfc_df)
coMethBetaDF_ls <- lapply(
  pfc_cgi_ls,
  function(x) pfc_df[which(CpGnames %in% x), ]
)
Sys.time() - a0
# 2.492514 min in serial
```

The first element is the data frame that matches the CpG sub-region above (with rows ordered by their location):
```{r inspect_subdata, eval=FALSE}
coMethBetaDF_ls$`chr10:100028320-100028499`[, 1:5]
```
```
           GSM1443251 GSM1443256 GSM1443263 GSM1443269 GSM1443270
cg04677227  0.1282135  0.1520103  0.1439823  0.1527586  0.1501033
cg07146435  0.3532991  0.3799506  0.3568279  0.3709329  0.3869927
cg11632906  0.2517267  0.3002839  0.2772760  0.3301750  0.3134411
```

### Create a New Cluster
If we wanted to perform a different calculation with the same data, we could use the exact same cluster as before, because these workers do not "expire" after use. However, because we need access to new data, we will create a new cluster of workers. It's often easier to create a new set of workers than try to modify the old workers.

Another condition to consider: fitting the linear mixed model on this data takes nearly 1Gb of ram, so take this into consideration when you create these clusters. If you already tapped all of your available RAM in the last computation, then make this cluster of workers with 1-2 fewer workers to account for this extra RAM you need. Because I had only used 52Gb of the 64Gb of RAM my machine has available, I could afford to perform the linear mixed model computation on a 12-worker cluster.
```{r parallel_createCluster2, eval=FALSE}
# 12 cores
clust <- makeCluster(12)
# Load coMethDMR for each worker
clusterEvalQ(cl = clust, library(coMethDMR))
# Load DMRcate for each worker
clusterEvalQ(cl = clust, library(DMRcate))
# Export the list of cluster-specific data frames and the response data
clusterExport(
  cl = clust,
  varlist = c("coMethBetaDF_ls", "pfcPheno_df")
)
```

### Linear Mixed Models in Parallel
Now we fit the models. Notice that this is considerably faster; even if you need to reduce the number of workers due to RAM constraints, this is still quite fast.
```{r parallel_lmm, eval=FALSE}
a1 <- Sys.time()
results_ls <- parLapply(
  cl = clust,
  coMethBetaDF_ls,
  fun = lmmTest,
  pheno_df = pfcPheno_df,
  contPheno_char = "stage",
  covariates_char = c("age.brain", "sex", "Mplate", "prop.neuron"),
  modelType = "randCoef",
  arrayType = "450k"
)
Sys.time() - a1
# 3.735125 min over 12 cores
```

A benefit to the **parallel** version is that if we want to repeat this computation with the simple linear mixed model, rather than the random coefficient version, we do not have to re-initialize the cluster.

### Post-computation Wrangling
Recall that the output for a single region is a one-row data frame, so we stack these data frames with `rbind()`.
```{r wrangle_lmmOut1, eval=FALSE}
res_cgi1_df <- do.call(rbind, results_ls)
res_cgi1_df$FDR <- p.adjust(res_cgi1_df$pValue, method = "fdr")
row.names(res_cgi1_df) <- NULL
```


# Using the **BiocParallel** Package

## Cluster Setup
For **BiocParallel**, the cluster setup is very easy.
```{r BiocParallel_setup, eval=FALSE}
library(BiocParallel)
snow_cl <- SnowParam(workers = 12, type = "SOCK")
```

## Finding the Co-Methylated Regions
While setting up the cluster is very easy, computing on it is more challenging.

### Creating the Wrapper Function
In the **parallel** cluster setup, we created the clusters, loaded the packages on them, and then exported the data we need to each. The **BiocParallel** cluster computing requires that we create a wrapper function to load the packages for each worker, but it does not require that we export any data to these workers.
```{r BiocParallel_wrapper, eval=FALSE}
worker_fun <- function(x, beta_mat){

  suppressPackageStartupMessages({
    library(coMethDMR)
    library(DMRcate)
  })


  CoMethSingleRegion(
    CpGs_char = x,
    betaMatrix = beta_mat,
    method = "pearson",
    arrayType = "450k",
    returnAllCpGs = FALSE
  )

}
```
This function requires us to understand exactly how the worker should act to complete your computation. Notice that we are calling the function we care about, `CoMethSingleRegion()` directly within this wrapper. Also note that we are loading the packages within this wrapper function. The `x` value will be the individual element of the list we are going to apply over, and the `beta_mat` is an argument placeholder for the `pfc_df` data.

### Executing the Wrapper Function on each Worker
```{r BiocParallel_bplapply, eval=FALSE}
a <- Sys.time()
coMethCpGsAllRegions_ls <- bplapply(
  # The list of regions to apply over
  X = closeByGenomicRegion_ls,
  # The worker function that calls CoMethSingleRegion()
  FUN = worker_fun,
  # Specify the cluster of workers
  BPPARAM = snow_cl,
  # Additional data passed through the worker function to CoMethSingleRegion()
  beta_mat = pfc_df
)
Sys.time() - a
# 14.83241 min over 12 cores
```
Note that this function doesn't appear to be as fast, but that's because this call to `bplapply()` loads the packages and exports the data for each worker as part of the computation. One of the main benefits here is that we don't have to export the data to the clusters manually. The cost here is that we can't re-use these clusters with the same data or packages loaded.

### Post-computational Wrangling
The results haven't changed, only the computing setup. We still have the same wrangling to do here.
```{r BiocParallel_wrangling, eval=FALSE}
pfc_cgi_rdrop0_4_ls <- unlist(
  lapply(coMethCpGsAllRegions_ls, `[[`, 2),
  recursive = FALSE
)
```

## Testing the Co-Methylated Regions
Once again, we have the results from `CoMethSingleRegion()` to pass along to a parallel version of `lmmTest()`. In this case, however, we do not have to create a new cluster with the updated data. The `bplapply()` function handles distributing the data for us, so we only need to keep track of the Snow-style cluster of workers we created above.

### Create the Linear Mixed Model Wrapper
As before, we need a worker function.
```{r BiocParallel_worker2, eval=FALSE}
worker2_fun <- function(cgi, beta_mat, pheno_mat){

  suppressPackageStartupMessages({
    library(coMethDMR)
    library(DMRcate)
  })

  coMethBetaDF <- beta_mat[which(rownames(beta_mat) %in% cgi), ]

  lmmTest(
    betaOne_df = coMethBetaDF,
    pheno_df = pheno_mat,
    contPheno_char = "stage",
    covariates_char = c("age.brain", "sex", "Mplate", "prop.neuron"),
    modelType = "randCoef",
    arrayType = "450k"
  )

}
```
This function is a touch more complicated than before, simply because we now have three arguments: the vector of co-methylated CpG locations, the data frame of methylation values, and the data frame of response and covariates.

### Apply the Wrapper
Once again, the code to execute this function in parallel is rather simple. The difficult part was creating the wrapper function correctly.
```{r BiocParallel_lmm, eval=FALSE}
a1 <- Sys.time()
results_ls <- bplapply(
  X = pfc_cgi_ls,
  FUN = worker2_fun,
  BPPARAM = snow_cl,
  beta_mat = pfc_df,
  pheno_mat = pfcPheno_df
)
Sys.time() - a1
# 6.800498 min over 12 cores
```
This execution is a little slower than it should be, but that is most likely due to the fact that the **parallel** version used a pre-computed list of data frames, while this version subsets the `pfc_df` data within the loop. For the previous version, we spent 2.5 minutes splitting the data frame into a list of co-methylated, region-specific data frames before executing the parallel code. This version performs that step within the worker function.

### Post-compute Wrangling
Because `lmmTest()` returns a one-row data frame, we still have to stack these results.
```{r BiocParallel_wangling2, eval=FALSE}
res_cgi1_df <- do.call(rbind, results_ls)
res_cgi1_df$FDR <- p.adjust(res_cgi1_df$pValue, method = "fdr")
row.names(res_cgi1_df) <- NULL
```


# Genome-Wide coMethDMR
Now that we have described how to cluster the co-methylated regions within one region type and test these clusters, we can expand this framework to include all region types.

```{r eval=FALSE}

# ANALYSIS FOR ALL REGIONS, EXCEPT CGIs

setwd("C:/Users/lxw391/Box Sync/METHODS-SHARED/METHOD-METHYL-GENE-TEST/coMethDMR_analysis_ADdata")

data.dir <- "C:/Users/lxw391/Box Sync/METHODS-SHARED/METHOD-METHYL-GENE-TEST/AD/DATA/"

####### first install package from gitlab

 # devtools::install_git(
 #     "https://gitlab.com/lissettegomez/coMethDMR.git",
 #     credentials = git2r::cred_user_pass("lilyAtGitlab", getPass::getPass())
 #  )

library(coMethDMR)

###########################################################

#  analysis of PFC data form Lunnon et al. (2014)
#  - using linear mixed model

###########################################################

## 1. import bmiq normalized data

pfc_df <- readRDS (paste0(data.dir, "PFCbetas.bmiq.RDS"))


# make pheno file
pheno <- readRDS (paste0(data.dir, "pheno.pfc.RDS"))

prop.neuron <-readRDS("pfc.prop.neurons.RDS")

pfcPheno_df <- merge ( pheno, prop.neuron, by.x="sample.id", by.y="row.names")

colnames(pfcPheno_df)[1] <- "Sample"

## 2. load closeby regions that are in CGIs -> compute co-methylated regions

test_ls <- readRDS(system.file(
        "extdata", "NSHORE3_200.RDS",
         package = 'coMethDMR', mustWork = TRUE
      ))

library(parallel)

clust <- makeCluster(22)
clusterEvalQ(cl = clust, library(coMethDMR))

#clusterExport(cl = clust, varlist = c("fun.dir" ) )

#clusterEvalQ ( cl= clust, source(paste0(fun.dir, "prediction.elastic.net.tryCatch.R")) )

regions_char <- c("NSHORE", "NSHELF", "SSHORE", "SSHELF",
                  "TSS1500", "TSS200", "UTR5", "EXON1", "GENEBODY",
                  "UTR3")

for (i in 1:10){

  out_ls <- CoMethAllRegions (
    betaMatrix = pfc_df,
    regionType = regions_char[i],
    arrayType = "450k",
    returnAllCpGs = FALSE
  )

  saveRDS (out_ls, paste0("coMethRegions_", regions_char[i], ".RDS"))

}

# ## 3. Test one region
#
# pfc_cgi_ls <- readRDS ("pfc_cgi_ls.RDS")
#
# cpgs_char <- GetCpGsInRegion(
#   regionName_char = "chr22:19709548-19709755",
#   arrayType = "450k"
# )
#
# coMethCpGs <- CoMethSingleRegion(CpGs_char = cpgs_char, betaMatrix = pfc_df)
#
# oneRegion_df <- pfc_df[coMethCpGs$CpGsSubregions[[1]] ,]
#
# lmmTest (betaOne_df = oneRegion_df,
#          pheno_df = pfcPheno_df,
#          contPheno_char = "stage",
#          covariates_char = c("age.brain", "sex", "Mplate", "prop.neuron"),
#          modelType = "randCoef",
#          arrayType = "450k")

## 4. test all regions

for (i in 1:10) {

  coMethRegions_ls <- readRDS(paste0("coMethRegions_", regions_char[i], ".RDS"))

  results_df <- lmmTestAllRegions(
                      beta_df = pfc_df,
                      region_ls = coMethRegions_ls$CpGsSubregions,
                      pheno_df = pfcPheno_df,
                      contPheno_char = "stage",
                      covariates_char = c("age.brain", "sex", "Mplate", "prop.neuron"),
                      modelType = "randCoef",
                      arrayType = "450k"
                    )

  AddFDRs (dat = results_df, fileName = paste0("coMethRegions_", regions_char[i]))

}

gc()
stopCluster(clust)



```


# Comments
Overall, these methods have similar computational and memory costs. Comparison:

- **Cluster Setup**:
    + `parallel::parLapply()` requires a two-step process. It takes time at the very beginning to initialize all the workers, then the cluster is available for as many uses as needed, so long as the data input remains constant.
    + The `BiocParallel::bplapply()` function initializes the pre-defined cluster setup and performs computation at the same time. This is much simpler to set up, but requires re-initialization before each new computational task.
- **Speed**: Overall, the computational costs are very similar between the two methods.
- **Executing Jobs**:
    + For simple problems, the **BiocParallel** version is much nicer. You make a single call to create the cluster, and `bplapply()` handles all the setup as part of the execution.
    + `parLapply()` requires the same type of setup no matter how easy or complicated the job.
    + Comparison: for easy jobs, this is a solid win for `bplapply()`; for more complicated parallel computing jobs, this is a solid win for `parLapply()`.

In summary, I think it is easier to write package code to use **parallel**. The package setup to allow users the **BiocParallel** option is more difficult. However, allowing users to create their own clusters with **BiocParallel** is a very strong bonus. Additionally, it will make the Bioconductor folks happy if we use their parallel routines. If you write code for your own use, I think that **parallel** is easier to use. If you write code for other people to sue from a package, **BiocParallel** will make your package easier for other people to use.
